Analyzing Violent Crime Statistics in Toronto Neighbourhoods

### 2024-01-25 outputs/paper/paper.qmd
Updated intro

While we read about rising and falling crime rates in the cities we live in, we don't know if these crime rates affect every person equally. It could be the case that these crime occurrences are distributed evenly throughout the city, where every person living in different parts of the city has the same probability of being a victim of a crime. However, it could also be the case that not every person will run into these crimes in their day to day life, and this probability will be heavily influenced by where they live. The aim of this paper is to explore if all neighbourhoods of Toronto follow the general trends of rise and fall in violent crimes, or are there variances between neighbourhoods.

@@

The data includes counts for individual occurence for each category of crime - assault, auto theft, break and entering, robbery, homicide, theft over (\$5000), shooting and firearm discharges. The data also includes the crime rate for each of these categories, calculated as crimes committed per 100,000 population (following the standard definition of crime rate set by Statistics Canada). These crime rates were calculated using population estimates for each neighborhood in 2023, which were provided by Environics Analytics. Both the individual crime occurences as well as crime rates have been given for the year 2014 to 2023, where each year gets its own column [@crimedata].

```{r}
#| echo: false
#| message: false
#| label: sample_assault_data
#| tbl-cap: Sample of assault count and assault rate arranged according to years
data_snippet <- cleaned_data[1:3, c("assault_2022", "assault_2023", "assault_rate_2014", "assault_rate_2015")]

@@

For our analysis, we will primarily be using vertical box plots to plot our data, as it gives us a clear view of how different crime rates are distributed amongst the neighbourhoods of Toronto.

The top whisker (slim black line) of our box plot represents the top 25% of our data, and the bottom whisker represents the bottom 25% of the data. The black line inside the box of the plots represents the median of the data. The top segment of the box represents the top 50% to top 75% of our data, whereas the bottom segment of the box represents the bottom 25% to the bottom 50% of our data.

@@

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-assault_rate_dist
#| fig-cap: Box plots of assault rate distribution in Toronto Neighbourhoods over the years
num_assault <- cleaned_data |> select(contains("assault_rate_2"))
headers <- as.character(2014:2023)
names(num_assault) <- headers
num_assault <- melt(num_assault)
ggplot(num_assault, aes(x=variable, y=value)) +
    geom_boxplot(outlier.shape = NA) +
    xlab("Year") +
    ylab("Assaults / 100,000 population") +
    ylim(0,300) +
    ggtitle(wrapper("Distribution of assault rates across all neighbourhoods per year", width=60)) +
    theme(plot.title = element_text(size=12, hjust = 0.5),
          axis.title = element_text(size=10))
```

@@

## Robbery Rates {#subsec-results-robbery}
```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-robbery_rate_dist
#| fig-cap: Box plots of robbery rate distribution in Toronto Neighbourhoods over the years
num_robbery <- cleaned_data |> select(contains("robbery_rate_2"))
headers <- as.character(2014:2023)
names(num_robbery) <- headers
num_robbery <- melt(num_robbery)
ggplot(num_robbery, aes(x=variable, y=value)) +
    geom_boxplot(outlier.shape = NA) +
    xlab("Year") +
    ylab("Robberies / 100,000 population") +
    ylim(0,60) +
    ggtitle(wrapper("Distribution of robbery rates across all neighbourhoods per year", width=60)) +
    theme(plot.title = element_text(size=12, hjust = 0.5),
          axis.title = element_text(size=10))
```
Similar to assaults, in @fig-robbery_rate_dist, we can see that throughout the years, robbery rates are fairly normally distributed and that there is large variance in amongst the neighbourhoods. Top 25% of the neighbourhoods have a robbbery rate as high as 50-60%, whereas the bottom 25% go as low as 0 robberies. Such a high variance is indicative of massive differences amongst neighbourhoods of Toronto - not all neighbourhoods experience the same level of robberies. Notice that this is where the similarities end.

@@

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-homicide_rate_dist
#| fig-cap: Homicide Rate distribution in Toronto Neighbourhoods over the years
num_homicide <- cleaned_data |> select(contains("homicide_rate_2"))
headers <- as.character(2014:2023)
names(num_homicide) <- headers
num_homicide <- melt(num_homicide)
ggplot(num_homicide, aes(x=variable, y=value)) +
    geom_boxplot(outlier.shape = NA) +
    xlab("Year") +
    ylab("Homicides / 100,000 population") +
    ylim(NA, 20) +
    ggtitle(wrapper("Distribution of homicide rates across all neighbourhoods per year", width=60)) +
    theme(plot.title = element_text(size=12, hjust = 0.5),
          axis.title = element_text(size=10))
```

@@

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-shooting_rate_dist
#| fig-cap: Box plots of shooting rate distribution in Toronto Neighbourhoods over the years
num_shooting <- cleaned_data |> select(contains("shooting_rate_2"))
headers <- as.character(2014:2023)
names(num_shooting) <- headers
num_shooting <- melt(num_shooting)
ggplot(num_shooting, aes(x=variable, y=value)) +
    geom_boxplot(outlier.shape = NA) +
    xlab("Year") +
    ylab("Shootings / 100,000 population") +
    ylim(0,12) +
    ggtitle(wrapper("Distribution of shooting rates across all neighbourhoods per year", width=60)) +
    theme(plot.title = element_text(size=12, hjust = 0.5),
          axis.title = element_text(size=10))
```

>>>>

Over the last five years, Toronto has seen a steady rise in homicides [@citynewshomicides], but at the same time gun violence has dropped 30% [@cbcgunviolence]. In 2022, the city saw a 15% increase in its violent crime index, which was more than three times the increase Canada saw nationally [@cp24crimejump]. While we read about rising and falling crime rates in the cities we live in, we don't know if these crime rates affect every person equally. It could be the case that these crime occurrences are distributed evenly throughout the city, where every person living in different parts of the city has the same probability of being a victim of a crime. However, it could also be the case that not every person will run into these crimes in their day to day life, and this probability will be heavily influenced by where they live. The aim of this paper is to explore if all neighbourhoods of Toronto follow the general trends of rise and fall in violent crimes, or are there variances between neighbourhoods.

@@

The data includes counts for individual occurence for each category of crime - assault, auto theft, break and entering, robbery, homicide, theft over (\$5000), shooting and firearm discharges. The data also includes the crime rate for each of these categories, calculated as crimes committed per 100,000 population (following the standard definition of crime rate set by Statistics Canada). These crime rates were calculated using population estimates for each neighborhood in 2023, which were provided by Environics Analytics. Both the individual crime occurences as well as crime rates have been given for the year 2014 to 2023, where each year gets its own column [@crimedata], as seen in @tbl-sample_assault_data.

```{r}
#| echo: false
#| message: false
#| label: tbl-sample_assault_data
#| tbl-cap: Sample of assault count and assault rate arranged according to years
data_snippet <- cleaned_data[1:3, c("assault_2022", "assault_2023", "assault_rate_2014", "assault_rate_2015")]

@@

@@

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-assault_rate_dist
#| fig-cap: Box plots of assault rate distribution in Toronto Neighbourhoods over the years. Median rate for each year is represented by horizontal lines inside the boxes.
num_assault <- cleaned_data |> select(contains("assault_rate_2"))
headers <- as.character(2014:2023)
names(num_assault) <- headers
num_assault <- melt(num_assault)
ggplot(num_assault, aes(x=variable, y=value)) +
    geom_boxplot(outlier.shape = NA) +
    xlab("Year") +
    ylab("Assaults / 100,000 population") +
    ylim(0,300) +
    ggtitle(wrapper("Distribution of assault rates across all neighbourhoods per year", width=60)) +
    theme(plot.title = element_text(size=12, hjust = 0.5),
          axis.title = element_text(size=10))
```

@@

## Robbery Rates {#subsec-results-robbery}
```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-robbery_rate_dist
#| fig-cap: Box plots of robbery rate distribution in Toronto Neighbourhoods over the years. Median rate for each year is represented by horizontal lines inside the boxes.
num_robbery <- cleaned_data |> select(contains("robbery_rate_2"))
headers <- as.character(2014:2023)
names(num_robbery) <- headers
num_robbery <- melt(num_robbery)
ggplot(num_robbery, aes(x=variable, y=value)) +
    geom_boxplot(outlier.shape = NA) +
    xlab("Year") +
    ylab("Robberies / 100,000 population") +
    ylim(0,60) +
    ggtitle(wrapper("Distribution of robbery rates across all neighbourhoods per year", width=60)) +
    theme(plot.title = element_text(size=12, hjust = 0.5),
          axis.title = element_text(size=10))
```
Similar to assaults, in @fig-robbery_rate_dist, we can see that throughout the years, robbery rates are fairly normally distributed and that there is large variance in amongst the neighbourhoods. Top 25% of the neighbourhoods have a robbbery rate as high as 50-60%, whereas the bottom 25% go as low as 0 robberies. Such a high variance is indicative of massive differences amongst neighbourhoods of Toronto - not all neighbourhoods experience the same level of robberies. Notice that this is where the similarities end.

@@

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-homicide_rate_dist
#| fig-cap: Box plots of homicide rate distribution in Toronto Neighbourhoods over the years. Median rate for each year is represented by horizontal lines inside the boxes.
num_homicide <- cleaned_data |> select(contains("homicide_rate_2"))
headers <- as.character(2014:2023)
names(num_homicide) <- headers
num_homicide <- melt(num_homicide)
ggplot(num_homicide, aes(x=variable, y=value)) +
    geom_boxplot(outlier.shape = NA) +
    xlab("Year") +
    ylab("Homicides / 100,000 population") +
    ylim(NA, 20) +
    ggtitle(wrapper("Distribution of homicide rates across all neighbourhoods per year", width=60)) +
    theme(plot.title = element_text(size=12, hjust = 0.5),
          axis.title = element_text(size=10))
```

@@

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-shooting_rate_dist
#| fig-cap: Box plots of shooting rate distribution in Toronto Neighbourhoods over the years. Median rate for each year is represented by horizontal lines inside the boxes.
num_shooting <- cleaned_data |> select(contains("shooting_rate_2"))
headers <- as.character(2014:2023)
names(num_shooting) <- headers
num_shooting <- melt(num_shooting)
ggplot(num_shooting, aes(x=variable, y=value)) +
    geom_boxplot(outlier.shape = NA) +
    xlab("Year") +
    ylab("Shootings / 100,000 population") +
    ylim(0,12) +
    ggtitle(wrapper("Distribution of shooting rates across all neighbourhoods per year", width=60)) +
    theme(plot.title = element_text(size=12, hjust = 0.5),
          axis.title = element_text(size=10))
```
